---
title: "Robust Regression on Baseball Cards"
output:
  pdf_document:
    number_sections: no
urlcolor: blue
---

In "A Comprehensive Survey of Loss Functions in Machine Learning", by Qi Wang, Yue Ma, Kun Zhao & Yingjie Tian, available at  https://link.springer.com/article/10.1007/s40745-020-00253-5, one alternative to Huber loss that is discussed is Log-cosh loss, defined as 

$$\rho(\mathcal{P}, u) = \sum_{u \in \mathcal{P}}\left( \log(\cosh(r_u)) \right)$$

where $r_u$ is the residual $y_u - \hat y_u$, and $\cosh$ is the hyperbolic cosine.

a) (2 marks) Plot `log(cosh(x))` and comment on its appropriateness as a loss function.
```{r}
x <- seq(-10, 10, length.out = 1000)
y <- log(cosh(x))
plot(x, y, type = "l", col = "blue", lwd = 2)

```
It is appropriate for a loss function since it is a convex function and clearly has a global minimum.

b) (1 mark) Write a function of log-cosh loss called `logcoshfn`.
$$\rho(\hat{\theta}) = \sum_{u \in \mathcal{P}} \log(\cosh(y_u-\alpha-\beta(x-\bar{x}))) ,\ \theta=(\alpha, \beta)$$

```{r}
cards = read.csv("data/df_cards.csv")

createlogcoshRho <- function(x, y) {
  ## local variable
  xbar <- mean(x)
  ## Return this function
  function(theta) { 
    alpha <- theta[1]
    beta <- theta[2]
    sum( log(cosh(y - alpha - beta * (x - xbar))) )
  }
}

logcoshfn = createlogcoshRho(cards$war, cards$n)
```

c) (3 marks) Find the derivative, write it **in LaTeX**

$$\frac{d}{dx}\left[ \log(\cosh(x))\right] = \frac{\sinh(x)}{\cosh(x)} = \tanh(x) $$

$$\rho(\hat{\theta}) = \sum_{u \in \mathcal{P}} \log(\cosh(y_u-\alpha-\beta(x-\bar{x}))) ,\ \ \theta=(\alpha, \beta)$$
$$g=\nabla \rho(\theta)=\left[\begin{array}{l}
\frac{d \rho}{d \alpha} \\
\frac{d \rho}{d \beta}
\end{array}\right]=\left[\begin{array}{l}
\sum_{u \in \rho}-\tanh \left(y_u-\alpha-\beta(x-\bar{x})\right) \\
\sum_{u \in \rho}-(x-\bar{x}) \tanh \left(y_u-\alpha-\beta(x-\bar{x})\right)
\end{array}\right]=\left[\begin{array}{l}
\sum_{u \in \rho}-\tanh \left(r_u\right) \\
\sum_{u \in \rho}-(x-\bar{x}) \tanh \left(r_u\right)
\end{array}\right]$$

d) (1 mark) Write a function of the derivative of log-cosh loss called `logcoshprimefn`.
```{r}
createlogcoshGradient <- function(x, y) {
  ## local variables
  xbar <- mean(x)
  ybar <- mean(y)
  function(theta) {
    alpha <- theta[1]
    beta <- theta[2]
    ru = y - alpha - beta*(x - xbar)
    -1*c( sum(tanh(ru)), 
          sum(tanh(ru)*(x-xbar)) )
  }
}

logcoshprimefn = createlogcoshGradient(cards$war, cards$n)
```

e) (5 marks) Use the `logcoshfn` and `logcoshprimefn` functions as `rho` and `gradient` in the `gradientdescent` function, and find a robust estimate of the number of cards published (`n`, y) as a function of the career wins above replacement (WAR) (`war` x) of a baseball player in `df_cards.csv`, and use it to fit a line to the data and plot it. *(Note: WAR is a measure of how much a player has contributed to their team's success, compared to a 'replacement' player who would be someone brought up from a league below. It's possible to have a WAR below zero, but in this dataset, those cases have been removed. In short, higher WAR means a better player.)*

Use $\hat\theta_0 = (7,2)$

```{r}
gradientDescent <- function(theta = 0,
      rhoFn, gradientFn, lineSearchFn, testConvergenceFn,
      maxIterations = 100,
      tolerance = 1E-6, relative = FALSE,
      lambdaStepsize = 0.01, lambdaMax = 0.5 ) {

  converged <- FALSE
  i <- 0

  while (!converged & i <= maxIterations) {
    g <- gradientFn(theta) ## gradient
    glength <-  sqrt(sum(g^2)) ## gradient direction
    if (glength > 0) d <- g /glength

    lambda <- lineSearchFn(theta, rhoFn, d,
                lambdaStepsize = lambdaStepsize, lambdaMax = lambdaMax)

    thetaNew <- theta - lambda * d
    converged <- testConvergenceFn(thetaNew, theta,
                                   tolerance = tolerance,
                                   relative = relative)
    theta <- thetaNew
    i <- i + 1
  }

  ## Return last value and whether converged or not
  list(theta = theta, converged = converged, iteration = i, fnValue = rhoFn(theta))
}
```

```{r}
### line searching could be done as a simple grid search
gridLineSearch <- function(theta, rhoFn, d,lambdaStepsize = 0.01, lambdaMax = 1) {
  ## grid of lambda values to search
  lambdas <- seq(from = 0, by = lambdaStepsize,  to = lambdaMax)
  ## line search
  rhoVals <- sapply(lambdas, function(lambda) {rhoFn(theta - lambda * d)})
  ## Return the lambda that gave the minimum
  lambdas[which.min(rhoVals)]
}
```


```{r}
### Where testCovergence might be (relative or absolute)
testConvergence <- function(thetaNew, thetaOld, tolerance = 1E-10, relative=FALSE) {
   sum(abs(thetaNew - thetaOld)) < if (relative) tolerance * sum(abs(thetaOld)) else tolerance
}


```


```{r}
gradientDescent(theta = c(7, 2), rhoFn = logcoshfn, gradientFn = logcoshprimefn, 
                lineSearchFn = gridLineSearch, testConvergenceFn = testConvergence, 
                lambdaStepsize = 0.001)


plot(cards$n, cards$war)
abline(7, 2)
abline(5.1895266, 0.1694879, col = "blue")
```

f) (1 mark) Check your work using the `optim` function with the same initial values for `par` and `rho` as the `fn`.
```{r}
optim(par=c(7, 2), fn = logcoshfn)
```
